---
title: "DISTPROF Exploratory Data Analysis"
format: html
---

# Data Wrangling

## Loading in the data

```{r}
setwd("/Users/biancaschutz/HERC_Sp25")
```

```{r}
load_distprof_data <- function(years) {
  distprof_dataframes <- list()
  for (i in 1:length(years)) {
    year <- years[[i]]
    path <- paste0("0_Datasets_csv/Data", year, "/District/clean_data/distprof_", year, "_clean.csv")
    df <- read.csv(path)
    distprof_dataframes[[i]] <- data.frame(df)
    names(distprof_dataframes)[[i]] <- paste0("distprof", year)
    cat("Loaded in DISTPROF for", year, "\n")
  }
  return(distprof_dataframes)
}

distprof_list <- load_distprof_data(2020:2023)
```

## Remove Columns with Too Many NAs

```{r}
library(dplyr)

calculate_missing_percentage <- function(df) {
  na_counts <- lapply(df, function(x) sum(is.na(x))/nrow(df))
  return(na_counts)
}

drop_columns <- function(df, threshold = .5) {
  percents <- calculate_missing_percentage(df)
  column_ref <- names(percents)
  percent_ref <- unname(unlist(percents))

  drop_cols <- data.frame(column = column_ref, percent_na = percent_ref) %>%
    filter(percent_na >= threshold) %>% pull(column)
  
  new_df <- df %>% select(-all_of(drop_cols))
  
  return(new_df)
}

# drop_columns(distprof_list[[1]])
```

```{r}
distprof_drop_cols <- lapply(distprof_list, drop_columns)
```

## Extract Relevant Columns

For DISTPROF, we are interested in the following data:
* African American Student Percent
* American Indian Student Percent
* Asian American Student Percent
* Hispanic Student Percent
* Pacific Islander Student Percent
* Two or More Races Student Percent
* White Student Percent
* Economically Disadvantaged Student Percent
* Special Education Student Percent
* Gifted & Talented Student Percent
* Immigrant Student Percent
* Bilingual & ESL Student Percent
* Teacher to Student Ratio
* All Students Count/number of students in district/enrollment

```{r}
distprof_selected <- list()
for (i in 1:length(years)) {
  distprof_yr <- distprof_drop_cols[[i]]
  
  year <- years[[i]]
  
  column_names <- c(paste0("District.", year, ".Student.Membership..", c("White.Percent", "African.American.Percent", "Hispanic.Percent", "American.Indian.Percent", "Asian.Percent", "Pacific.Islander.Percent", "Two.or.More.Races.Percent", "Econ.Disadv.Percent", "Bilingual.ESL.Percent", "Gifted...Talented.Percent", "Immigrant.Percent", "Special.Ed.Percent", "All.Students.Count")), paste0("District.", year, ".Staff..Teacher.Student.Ratio"), "DISTRICT", "DISTNAME")
  
  distprof_selected[[i]] <- distprof_yr[,column_names]
  names(distprof_selected)[[i]] <- paste0("distprof", year)
}
```

## Rbind dataframes

```{r}
# new column names
for (i in 1:length(distprof_selected)) {
  column_names_no_year <- gsub("_+", "_", gsub("\\.{2}", "_",  
                             gsub("\\.", "_", 
                                  gsub("District\\.\\d{4}\\.", "",
                                       colnames(distprof_selected[[i]])))))

  colnames(distprof_selected[[i]]) <- column_names_no_year
  
  distprof_selected[[i]]$Year <- years[[i]]
}
```

```{r}
distprof <- do.call(rbind, distprof_selected)
rownames(distprof) <- NULL
```

First change colnames so they're the same
# Exploratory Data Analysis

## African American Student Percent

Takeaway: the schools with the largest percentage of Black students are college preparatory schools or academies (likely private schools).

```{r, fig.width = 10, fig.height = 5}
library(ggplot2)
distprof2023 <- distprof_selected[[4]]

top10_black_student_pct <- distprof2023 %>% arrange(desc(Student_Membership_African_American_Percent)) %>% slice_head(n = 10) %>% select(DISTNAME, Student_Membership_African_American_Percent)

black_student_pct <- ggplot(top10_black_student_pct, aes(x = DISTNAME, y = Student_Membership_African_American_Percent)) +
  geom_col(fill = 'green4') +
  scale_x_discrete(labels = scales::label_wrap(5)) +
  theme(legend.position = 'none') +
  labs(title = "10 Highest African American Student Populations in Texas School Districts, 2023",
       x = "District Name",
       y = "African American Student Percent")

black_student_pct
```
Next, let's look at the 

```{r}
# Insert code loading in DREF here

# then join by district ID and year, then filter to eliminate private/prep schools. 
```